---
title: "Tree-structured stick breaking process"
author: "Ke Yuan"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: null
    css: wch_style.css
    toc: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo = FALSE}
# wch's setting 
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The TSSB prior 

To illustrate the TSSB prior, we load the `bitphyloR` package and initiate a `Node` object. 
```{r}
library(bitphyloR)
n0 <- Node$new()
```
The `Node` object is the basic class representing components in the mixture. 
```{r}
n0
```

Now, let's initiate a `TSSB` object. In this case, 
```{r}
fixedSeed <- 9
set.seed(fixedSeed)
tssb <- TSSB$new(n0, data = matrix(rnorm(50),50,1), dpAlpha = 1, dpGamma = 1,
                 dpLambda = 1)
tssb
```

The tssb object generates an initial configuration of weights and tree.
```{r}
res <- tssb$GetMixture()
barplot(res$weight)
```


```{r}
res2 <- tssb$ConvertTssbToIgraph()
g <- res2$g
plot.igraph(g, layout = layout.reingold.tilford(g), 
     vertex.label = get.vertex.attribute(g, name = "size"),
     vertex.size = 15, edge.arrow.size = 0.5)
```

There are several empty leaf nodes. They can be removed by the `CullTree` function 
```{r}
tssb$CullTree()
res3 <- tssb$ConvertTssbToIgraph()
g <- res3$g
plot.igraph(g, layout = layout.reingold.tilford(g), 
     vertex.label = get.vertex.attribute(g, name = "size"),
     vertex.size = 15, edge.arrow.size = 0.5)
```

## Inference via MCMC

Let's test a simple dataset. 
```{r}

set.seed(fixedSeed)
m <- 200
dims <- 2
testData <- rbind(matrix(0.051*rnorm(m)+0.4,m/2,dims),
                  matrix( 0.01*rnorm(m)+0.6,m/2,dims))
plot(testData[,1], testData[,2])
```

Get an estimate of hyperparameters  
```{r}
empCov <- cov(testData)
priorSigmaScale = empCov + diag(rep(1e-6,dims), dims)
priorDriftScale = priorSigmaScale
```

We are going fit the data with a tree-structured normal mixture model. To do that, initiate a Normal node object
```{r}
n0 <- Normal$new(priorSigmaScale = priorSigmaScale, priorDriftScale = priorDriftScale)
n0
```

Initiate a TssbMCMC node object for MCMC functions. The initialiser also generates a guess of the tree structure and mass distribution based on tssb parameters.
```{r}
tssbMCMC <- TssbMCMC$new(n0, data = testData, dpAlpha = 1, dpGamma = 1, dpLambda = 1)
g0 <- tssbMCMC$ConvertTssbToIgraph()$g
plot.igraph(g0, layout = layout.reingold.tilford(g0), 
     vertex.label = get.vertex.attribute(g0, name = "size"),
     vertex.size = 15, edge.arrow.size = 0.5)
```

Sample the assignment for each point to clusters
```{r}
par(mfrow = c(1, 2))
plot.igraph(g0, layout = layout.reingold.tilford(g0), 
     vertex.label = get.vertex.attribute(g0, name = "size"),
     vertex.size = 15, edge.arrow.size = 0.5)
tssbMCMC$ResampleAssignments()
g1 <- tssbMCMC$ConvertTssbToIgraph()$g
plot.igraph(g1, layout = layout.reingold.tilford(g1), 
     vertex.label = get.vertex.attribute(g1, name = "size"),
     vertex.size = 15, edge.arrow.size = 0.5)
```

Remove empty path
```{r}
par(mfrow = c(1, 2))
plot.igraph(g1, layout = layout.reingold.tilford(g1), 
     vertex.label = get.vertex.attribute(g1, name = "size"),
     vertex.size = 15, edge.arrow.size = 0.5)
tssbMCMC$CullTree()
g2 <- tssbMCMC$ConvertTssbToIgraph()$g
plot.igraph(g2, layout = layout.reingold.tilford(g2), 
     vertex.label = get.vertex.attribute(g2, name = "size"),
     vertex.size = 15, edge.arrow.size = 0.5)
```

Sample cluster parameters
```{r}
tssbMCMC$ResampleNodeParameters()
```

Sample cluster hyperparameters, normally parameters of the transition. Note that this function can only be called from the root node object
```{r}
n0$ResampleHyperParams()
```

Sample sticks according to the new configuration
```{r}
par(mfrow = c(1, 2))
tssbMCMC$ResampleSticks()
res <- tssbMCMC$GetMixture()
barplot(res$weight)
sum(res$weight)
```

```{r}
tssbMCMC$ResampleStickOrders()
res <- tssbMCMC$GetMixture()
barplot(res$weight)
sum(res$weight)
```

Compare the trees after 10 iterations
```{r}
par(mfrow = c(1, 2))
g <- tssbMCMC$ConvertTssbToIgraph()$g
plot.igraph(g, layout = layout.reingold.tilford(g), 
     vertex.label = get.vertex.attribute(g, name = "size"),
     vertex.size = 15, edge.arrow.size = 0.5)

trace <- RunNormal(data = testData, numOfMCMC = 10, tssb = tssbMCMC)
g <- tssbMCMC$ConvertTssbToIgraph()$g
plot.igraph(g, layout = layout.reingold.tilford(g), 
     vertex.label = get.vertex.attribute(g, name = "size"),
     vertex.size = 15, edge.arrow.size = 0.5)
```

Now, let's check a longer run,
```{r, cache = T}
traces <- RunNormal(data = testData, numOfMCMC = 1000, tssb = tssbMCMC)
```

Plot the marginal likelihood trace and its autocorrelation function
```{r}
par(mfrow = c(1, 2))
plot(traces$likelihood, type = "l")
acf(traces$likelihood, lag.max = 200)
```



